Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-21 paper.Rnw
Integrated Mallar's comments up to methods

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} (
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
\multiatlas approaches. Thus, our method acts as an accurate \multiatlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 60 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb{} when applied to
first episode psychosis and Alzheimer's disease populations, and \mb{}
segmentations are compared with existing automated methods (\fsl{}, \freesurfer{},
\maper{}) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo % TODO: update
cross-validation on 60 subjects relative to manual segmentations,
and shows significantly lower variability in DSC than \multiatlas segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subfields with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt)
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.\\ \textbf{Conclusion:} We demonstrate
that \mb{} produces accurate hippocampal segmentations using only 5 atlases over
different hippocampal definitions, disease populations, and acquisition types,
as well as showing that accurate identification of the hippocampal subfields is
possible.
\end{abstract}

@@

The gold standard for neuroanatomical segmentation is manual delineation by an
expert human rater.  However, with the availability of increasingly large MRI
datasets, the time and expertise required for manual segmentation becomes
prohibitive \citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. This
effort is complicated by the fact that there is significant variation between
segmentation protocols with respect to specific anatomical boundaries of the
hippocampus \citep{Geuze2004} and this has led to efforts to create an unified
hippocampal segmentation protocol \citep{Jack2011}.  In addition, there is
controversy over the appropriate manual segmentation protocol to use in a
particular imaging study \citep{Nestor2012}.  Thus, a segmentation algorithm
that can easily adapt to different manual segmentation definitions would be of
significant benefit to the neuroimaging community.

Automated segmentation techniques that are reliable, objective, and reproducible
can be considered complementary to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
move them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to error in the estimated nonlinear transformation itself, partial
volume effects in label resampling, and irreconcilable differences between the
neuroanatomy represented within the atlas and target images.

@@

More recently, many groups have used multiple atlases to improve overall
segmentation accuracy (i.e. \multiatlas segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm (Simultaneous Truth And Performance Level
Estimation; \citet{Warfield2004}) that computes a probabilistic segmentation
using an expectation-maximization framework from an set of competing
segmentations, or others where a subset of segmentations can be estimated using
metrics such as the sum of squared differences in the regions of interest to be
segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{MallarChakravarty2012}.
The novelty of this manuscript is the extension of our \multiatlas methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

Of special relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing \multiatlas segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic \multiatlas strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using \multiatlas techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs{} segmentations. %TODO: which segmentations were used? SNT?

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer{} package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and
specific landmarks, the authors demonstrated accurate parcellation of
hippocampal subfields in unlabelled MRI volumes.

In this paper we have describe a thorough validation of the \mb{} algorithm for
segmentation of the hippocampus and its subfields. First, we address the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases \citep{MallarChakravarty2012} by performing a multi-fold experiment
using a subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI)
dataset over a range of parameters. We then perform a leave-one-out
validation to determine if hippocampal subfields can be accurately identified,
using the best parameters discovered in the first experiment.  To ensure that we
have not overfit our parameters to the aging or neurodegenerative brain, we also
apply \mb{} to a dataset of individuals suffering from first episode psychosis.
Finally, we validat our algorithm using all of the data available in the
\adnidataset{} sample and compare our segmentations to other popular
segmentation algorithms.

@@

The simplest form of \multiatlas segmentation, which we call {\em basic 
\multiatlas segmentation}, involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. The basic \multiatlas segmentation method
is described in detail in other publications
\citep{Collins2010,Heckemann2011,Aljabar2009}. When only a single atlas is used,
basic \multiatlas segmentation degenerates into model-based segmentation: labels
are propagated from the atlas to a target, and no label fusion is needed.

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic \multiatlas segmentation and \mb{}
  segmentation \label{fig:MAGeT}}
\end{figure}

@@

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings explored
are: atlas library size (1-9), template library size (1-20), registration method
(\ants{} or \animal{}, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  In each validation round, both a
\mb{} and \multiatlas segmentation is produced.  A total of $10 \times 60 \times
9 \times 20 \times 2 \times 3 =$ \Sexpr{10*60*9*20*2*3} validation rounds
were conducted and resulting segmentations analysed.

@@

  \item[Cross-correlation Weighted Majority Vote]
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{MallarChakravarty2012}.  The number of top ranked template library image labels
  is a configurable parameter and displayed as the size of the template library
  in the rest of the paper.

@@

The labels produced by \mb{} and \multiatlas segmentation are compared to the
manual labels using the Dice similarity coefficient, and the recorded value for
each subject at each parameter setting explored in this experiment is the
average over ten validation rounds.

Additionally, the sensitivity of \mb and \multiatlas to atlas and template
library composition is evaluated by comparing the variability in Dice scores
over all validation rounds at fixed parameter settings.  This is achieved by
first computing the variance of DSC scores in each block of ten validation
rounds per subject. The distribution of these statistics across all subjects is
then compared between \mb{} and \multiatlas using a Student's t-test. A
significant difference between distributions is taken to show either a larger or
smaller level of variability between methods.

@@

To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, \mb{} was applied to large dataset
from the ADNI project and the resulting segmentations were compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated
hippocampal segmentations (SNT) provided by ADNI.

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
SNT, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).  Clinical and
demographic data for the entire \adnidataset{} dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

Semi-automated segmentations of the left and right whole hippocampi are made
available with a subset of ADNI images \citep{Hsu2002}. These labels have been
generated using the SNT tool from Medtronic Surgical Navigation Technologies,
Louisville, CO (see Supplementary Materials for detailed discussion of the
segmentation process used).

@@

One author (MP) performed visual quality inspection for \mb{} and \freesurfer{} 
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (SNT, \maper{}, \freesurfer{}, \fsl{}, and \mb{}) and also
passing quality control inspection were included in the analysis.

\paragraph{Evaluation method}
As in Experiment 3, the SNT and Winterburn hippocampal segmentation protocols
differ in the neuroanatomical features delineated, and so we assessed \mb{} by
the degree of (Pearson) correlation of average hippocampal volume across
subjects. We also computed the correlation in hippocampal volume between
existing, established automated segmentation methods -- \fsl{}, \freesurfer{},
and \maper{}, and SNT semi-automated segmentations. Additionally, we evaluate
the volume-related fixed and proportional biases in all segmentation methods
using Bland-Altman plots \citep{Bland1986}.

@@

In this experiment we conduct 10 rounds of \mb{} and \multiatlas segmentation of
of 60 subjects using atlas and template library sizes varying from 1-9 and 1-20
images respectively, two registration algorithms (\ants{} or \animal{}), and three
label fusion techniques (unweighted, cross-correlation, and normalised mutual
information weighted majority vote).  Hippocampal \mb{}-based segmentations using
both \animal{} and \ants{} registration algorithms demonstrate good overlap with
manual segmentations (maximum mean DSC of \Sexpr{round(mb.best.kappa, digits=2)}
when using 9 atlases, 19 templates, ANTS registration, and majority vote label
fusion); Figure \ref{fig:ADNI1-xval-k-mean}).  Qualitatively, both \animal{} and
\ants{}-based segmentations demonstrate trend overlap accuracy that increases with
the size of atlas library and template library. Improvement in accuracy plateaus
with template libraries larger than ten images.

The use of \mb{} with \ants{} registration shows a pronounced increase
in segmentation accuracy over \mb{} with \animal{} registration, across all
other variable settings we tested. Additionally, by itself, using a weighted
voting strategy did not significantly improve segmentation accuracy, contrary to
the findings of \citet{Aljabar2009} in basic \multiatlas segmentation. Given
these findings, in the remainder of this section only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

With an increasing number of templates, \mb{} shows improvement over
\multiatlas-based segmentation in overlap accuracy when using the same number of
atlases and voting method (Figure \ref{fig:ADNI1-xval-k-diff}). The two methods
converge in accuracy when using seven atlases.  Peak improvement in \mb{} accuracy
(~0.02 DSC) is found when one atlas is used with a template library of 19
images.

In addition to an improvement in accuracy over \multiatlas-based segmentation,
\mb{} also shows a decrease in the variability of segmentation accuracy 
(Figure \ref{ADNI1-xval-variability}).  The size of template library 
necessary to reach a decrease ($p < 0.05$) in variance and standard deviation
grows with the size of atlas library used.  A template library of 19 images is
sufficient to show significant decrease in variance and standard deviation for
3-7 atlases.

@@

Based on the results from Experiment 1, in this experiment \mb{} was configured
with a template library of 21 randomly chosen subject images (7 from each
disease class) and majority vote label fusion.  The entire \adnidataset{} dataset
was segmented by \mb{}, and the resulting volumes compared with those obtained by
expert semi-automated segmentation (SNT), and three other automated segmentation
techniques: \maper{}, \freesurfer{}, and \fsl{}. Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection.  A total of \Sexpr{length(means.complete$RID)} images are
included in the following analysis, having met quality criteria and having
segmentations from every method.

We found a close relationship in total bilateral hippocampal volume between all
methods and the SNT semi-automated label volumes (Figure
\ref{fig:ADNI-volumes-plot}).  Volumes are correlated with Pearson $r > 0.78$
for all methods across disease categories.  Within disease categories (Figure
\ref{fig:ADNI-volumes-boxplot}), \mb{} is consistently well correlated to manual
volumes (Pearson $r > 0.85$), but appears to slightly over-estimate the volume
of the AD hippocampus.

>>>>

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that \multiatlas{} approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb{} 
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
\multiatlas{} approaches. Thus, our method acts as an accurate \multiatlas{}
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb{} works by propagating atlas segmentations to a template
library, formed from a subset of target images, via transformations estimated by
non-linear image registration. The resultant segmentations are then propagated
to each target image and fused using a label fusion method.  To explore the
effect of atlases and templates library size, registation and label
fusion method, we conduct a 10-fold Monte Carlo cross-validation of \mb{} of whole
hippocampal segmentation on 60 ADNI subjects, manually segmented using the
\prsnr{}-protocol, over a range of parameter settings. Using the best settings
found, we then conduct a leave-one-out cross-validation (LOOCV) of hippocampal
subfield segmentation using five high-resolution manually segmented atlases.
Two final experiments assess \mb{} when applied to first episode psychosis and
Alzheimer's disease populations, and \mb{} segmentations are compared with
existing automated methods (\fsl{}, \freesurfer{}, \maper{}) and biases are
explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb{} achieves a
mean Dice's Similarity Coefficient (DSC) of 0.88 (to \psegs{}) over 10-folds of
Monte Carlo % TODO: update cross-validation on 60 subjects, and shows
significantly lower variability in DSC than \multiatlas{} segmentation.
In a LOOCV, \mb{} reproduces the volumes of the cornu ammonis (CA) 1;
CA4/dentate % TODO: update
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subfields with a percent error in volume that is
at or lower than that produced by image resampling. \mb{} produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl{} and \freesurfer{}, \mb{} shows much smaller
fixed volume bias (within $250 mm^3$ on average) to semi-automated (\snt{})
segmentations available from ADNI, as well as a conservative, rather than
exaggerated, proportional volume bias.\\ \textbf{Conclusion:} We demonstrate
that \mb{} produces accurate hippocampal segmentations using only 5 atlases over
different hippocampal definitions, disease populations, and acquisition types,
as well as showing that accurate identification of the hippocampal subfields is
possible.
\end{abstract}

@@

The gold standard for neuroanatomical segmentation is manual delineation by an
expert human rater.  However, with the availability of increasingly large MRI
datasets, the time and expertise required for manual segmentation becomes
prohibitive \citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. This
effort is complicated by the fact that there is significant variation between
segmentation protocols with respect to specific anatomical boundaries of the
hippocampus \citep{Geuze2004} and this has led to efforts to create an unified
hippocampal segmentation protocol \citep{Jack2011,Boccardi2013,Boccardi2013a}.
In addition, there is controversy over the appropriate manual segmentation
protocol to use in a particular imaging study \citep{Nestor2012}. Thus, a
segmentation algorithm that can easily adapt to different manual segmentation
definitions would be of significant benefit to the neuroimaging community.

Automated segmentation techniques that are reliable, objective, and reproducible
can be considered complementary to manual segmentation. In the case of classical
model-based segmentation methods \citep{Haller1997,Csernansky1998}, an MRI atlas
that was previously manually labelled by an expert rater is matched to target
images using nonlinear registration methods. The resulting nonlinear
transformation is applied to the manual labels (i.e. {\em label propagation}) to
warp them into the target image space. While this methodology has been used
successfully in several contexts
\citep{Chakravarty2008,Chakravarty2009,Collins1995,Haller1997}, it is limited in
accuracy due to error in the estimated nonlinear transformation itself, partial
volume effects in label resampling, and irreconcilable differences between the
neuroanatomy represented within the atlas and target images.

@@

More recently, many groups have used multiple atlases to improve overall
segmentation accuracy (i.e. \multiatlas{} segmentation) over model-based approaches
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2010,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used to
merge these labels into the definitive segmentation for the target. In addition,
weighted voting procedures that use {\em atlas selection} techniques are often
used to exclude atlases from label fusion that are dissimilar to a target image
in order to reduce error from unrepresentative anatomy \citep{Aljabar2009}.
This involves the selection of a subset of atlases using a similarity metric
such as cross-correlation \citep{Aljabar2009} or normalized mutual information.
Such selection has the added benefit of significantly reducing the number of
nonlinear registrations. For example \citet{Collins2010} demonstrated that only
14 atlases, selected based on highest similarity between medial temporal lobe
neuroanatomy as evaluated by normalized mutual information \citep{Studholme1999}
from a library of 80 atlases, were required to achieve accurate segmentations of
the hippocampus. Additionally, several methods have been explored for label
fusion, including the STAPLE algorithm (Simultaneous Truth And Performance Level
Estimation; \citet{Warfield2004}) that computes a probabilistic segmentation
using an expectation-maximization framework from an set of competing
segmentations, or others where a subset of segmentations can be estimated using
metrics such as the sum of squared differences in the regions of interest to be
segmented \citep{Coupe2011}.

However, many of these methods require significant investment of time and
resources for the creation of the atlas library ranging between 30
\citep{Heckemann2006} and 80 \citep{Collins2010} manually segmented atlases.
This strategy has the main drawback of being inflexible as it does not easily
accommodate varying the definition of the hippocampal anatomy (such as the
commonly used heuristic of subdividing the hippocampus into head, body, and tail
\citep{Poppenk2011,Pruessner2000}). Furthermore, none of these methods have
demonstrated sufficient flexibility to accommodate atlases that are somehow
exceptional such as those derived from serial histological data
\citep{Chakravarty2006,Yelnik2007} or high-resolution MRI data that enables
robust identification of hippocampal subfields
\citep{Winterburn2013,Yushkevich2009,Mueller2009,VanLeemput2009,Wisse2012}. Due
to the recent availability of the latter, there has been increased interest in
the use of probabilistic methods for the identification of the hippocampal
subfields on standard T1-weighted images. Our group recently demonstrated that
through use of an intermediary automated segmentation stage, robust and accurate
segmentation of the striatum, pallidum, and thalamus using a single atlas
derived from serial histological data is possible \citep{Chakravarty2013}.
The novelty of this manuscript is the extension of our \multiatlas{} methodology
to the hippocampus using more than a single input atlas, while simultaneously
limiting the number of inputs used during segmentation, and demonstrating that
accurate identification of the hippocampal subfields is indeed possible using
this methodology.

Of particular relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing \multiatlas{} segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic \multiatlas{} strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using \multiatlas{} techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs{} segmentations. %TODO: which segmentations were used? SNT?

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer{} package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were labelled on highly-anistropic
($0.380mm$ in-plane resolution, $0.8mm$ coronal slice thickness) T2-weighted MR
images acquired from post-mortem medial temporal lobe samples. Using nonlinear
registration guided by manually derived hippocampus masks and specific
landmarks, the authors demonstrate accurate parcellation of hippocampal
subfields in unlabelled MRI volumes.

In this paper we have describe a thorough validation of the \mb{} algorithm for
segmentation of the hippocampus and its subfields. First, we address the
feasibility of whole hippocampus segmentation with a limited number of input
atlases \citep{Chakravarty2013} by performing a multi-fold validation experiment using a
subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset
manually segmented using the \prsnr{}-protocol, over a
range of atlas and template library sizes, registration and label fusion
methods. We then perform a leave-one-out validation to determine if
hippocampal subfields can be accurately identified, using the best parameters
discovered in the first experiment.  To ensure that we have not overfit our
parameters to the aging or neurodegenerative brain, we also apply \mb{} to a
dataset of individuals suffering from first episode psychosis.  Finally, we
validate our algorithm using all of the data available in the \adnidataset{}
sample and compare our segmentations to other popular segmentation algorithms.

@@

The simplest form of \multiatlas{} segmentation, which we call {\em basic 
\multiatlas{} segmentation}, involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. The basic \multiatlas{} segmentation method
is described in detail in other publications
\citep{Collins2010,Heckemann2011,Aljabar2009}. When only a single atlas is used,
basic \multiatlas{} segmentation degenerates into model-based segmentation: labels
are propagated from the atlas to a target, and no label fusion is needed.

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=\textwidth]{figure/MA-MAGeTBrain-Schematic}
  \end{centering}
  \caption{A schematic illustration of basic \multiatlas{} segmentation and \mb{}
  segmentation \label{fig:MAGeT}}
\end{figure}

@@

A total of ten validation rounds were performed on each subject in the dataset,
over each combination of parameter settings. The parameter settings explored
are: atlas library size (1-9), template library size (1-20), registration method
(\ants{} or \animal{}, described below), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote, described below).  In each validation round, both a
\mb{} and \multiatlas{} segmentation is produced.  A total of $10 \times 60 \times
9 \times 20 \times 2 \times 3 =$ \Sexpr{10*60*9*20*2*3} validation rounds
were conducted and resulting segmentations analysed.

@@

  \item[Cross-correlation Weighted Majority Vote]
  An optimal combination of targets from the template library has previously been
  shown to improve segmentation accuracy \citep{Aljabar2009,Collins2010}.  In this
  method, each template library image is ranked in similarity to each unlabelled 
  image by the normalized cross-correlation (CC) of image intensities after linear
  registration, over a region of interest (ROI) generously encompassing the 
  hippocampus.  Only the top ranked template library image labels are used in a
  voxel-wise majority vote. The ROI is heuristically defined as the extent of all
  atlas labels after linear registration to the template, dilated by three voxels
  \citep{Chakravarty2013}.  The number of top ranked template library image labels
  is a configurable parameter and displayed as the size of the template library
  in the rest of the paper.

@@

The labels produced by \mb{} and \multiatlas{} segmentation are compared to the
manual labels using the Dice similarity coefficient, and the recorded value for
each subject at each parameter setting explored in this experiment is the
average over ten validation rounds.

Additionally, the sensitivity of \mb{} and \multiatlas{} to atlas and template
library composition is evaluated by comparing the variability in Dice scores
over all validation rounds at fixed parameter settings.  This is achieved by
first computing the variance of DSC scores in each block of ten validation
rounds per subject. The distribution of these statistics across all subjects is
then compared between \mb{} and \multiatlas{} using a Student's t-test. A
significant difference between distributions is taken to show either a larger or
smaller level of variability between methods.

@@

To validate \mb{} segmentation quality with respect to other established
automated hippocampal segmentation methods, \mb{} was applied to large dataset
from the ADNI project and the resulting segmentations were compared to those
produced by \freesurfer{}, \fsl{}, \maper{}, as well as semi-automated
hippocampal segmentations (\snt{}) provided by ADNI.

\paragraph{ADNI1 dataset revisited} 
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper, \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).  Clinical and
demographic data for the entire \adnidataset{} dataset are shown in Table
\ref{tab:ADNI1-Dataset-Demographics}.

Semi-automated segmentations of the left and right whole hippocampi are made
available with a subset of ADNI images \citep{Hsu2002}. These labels have been
generated using the \snt{} tool from Medtronic Surgical Navigation Technologies,
Louisville, CO (see Supplementary Materials for detailed discussion of the
segmentation process used).

@@

One author (MP) performed visual quality inspection for \mb{} and \freesurfer{} 
segmentations using similar quality control guidelines (if either hippocampus
was under or over segmented by $10mm$ or greater in three or more slices then
the segmentation did not pass).  Only images meeting the conditions of having
segmentations from all methods (\snt{}, \maper{}, \freesurfer{}, \fsl{}, and \mb{}) and also
passing quality control inspection were included in the analysis.

\paragraph{Evaluation method}
As in Experiment 3, the \snt{} and Winterburn hippocampal segmentation protocols
differ in the neuroanatomical features delineated, and so we assessed \mb{} by
the degree of (Pearson) correlation of average hippocampal volume across
subjects. We also computed the correlation in hippocampal volume between
existing, established automated segmentation methods -- \fsl{}, \freesurfer{},
and \maper{}, and \snt{} semi-automated segmentations. Additionally, we evaluate
the volume-related fixed and proportional biases in all segmentation methods
using Bland-Altman plots \citep{Bland1986}.

@@

In this experiment we conduct 10 rounds of \mb{} and \multiatlas{} segmentation of
of 60 subjects using atlas and template library sizes varying from 1-9 and 1-20
images respectively, two registration algorithms (\ants{} or \animal{}), and three
label fusion techniques (unweighted, cross-correlation, and normalised mutual
information weighted majority vote).  Hippocampal \mb{}-based segmentations using
both \animal{} and \ants{} registration algorithms demonstrate good overlap with
manual segmentations (maximum mean DSC of \Sexpr{round(mb.best.kappa, digits=2)}
when using 9 atlases, 19 templates, ANTS registration, and majority vote label
fusion); Figure \ref{fig:ADNI1-xval-k-mean}).  Qualitatively, both \animal{} and
\ants{}-based segmentations demonstrate trend overlap accuracy that increases with
the size of atlas library and template library. Improvement in accuracy plateaus
with template libraries larger than ten images.

The use of \mb{} with \ants{} registration shows a pronounced increase
in segmentation accuracy over \mb{} with \animal{} registration, across all
other variable settings we tested. Additionally, by itself, using a weighted
voting strategy did not significantly improve segmentation accuracy, contrary to
the findings of \citet{Aljabar2009} in basic \multiatlas{} segmentation. Given
these findings, in the remainder of this section only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

With an increasing number of templates, \mb{} shows improvement over
\multiatlas{}-based segmentation in overlap accuracy when using the same number of
atlases and voting method (Figure \ref{fig:ADNI1-xval-k-diff}). The two methods
converge in accuracy when using seven atlases.  Peak improvement in \mb{} accuracy
(~0.02 DSC) is found when one atlas is used with a template library of 19
images.

In addition to an improvement in accuracy over \multiatlas{}-based segmentation,
\mb{} also shows a decrease in the variability of segmentation accuracy 
(Figure \ref{ADNI1-xval-variability}).  The size of template library 
necessary to reach a decrease ($p < 0.05$) in variance and standard deviation
grows with the size of atlas library used.  A template library of 19 images is
sufficient to show significant decrease in variance and standard deviation for
3-7 atlases.

@@

Based on the results from Experiment 1, in this experiment \mb{} was configured
with a template library of 21 randomly chosen subject images (7 from each
disease class) and majority vote label fusion.  The entire \adnidataset{} dataset
was segmented by \mb{}, and the resulting volumes compared with those obtained by
expert semi-automated segmentation (\snt{}), and three other automated segmentation
techniques: \maper{}, \freesurfer{}, and \fsl{}. Table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals} shows the total count of
segmentations available, including a count of those which have failed a quality
control inspection.  A total of \Sexpr{length(means.complete$RID)} images are
included in the following analysis, having met quality criteria and having
segmentations from every method.

We found a close relationship in total bilateral hippocampal volume between all
methods and the \snt{} semi-automated label volumes (Figure
\ref{fig:ADNI-volumes-plot}).  Volumes are correlated with Pearson $r > 0.78$
for all methods across disease categories.  Within disease categories (Figure
\ref{fig:ADNI-volumes-boxplot}), \mb{} is consistently well correlated to manual
volumes (Pearson $r > 0.85$), but appears to slightly over-estimate the volume
of the AD hippocampus.
